R Notebook

### 2022-04-12 src/main/r/biomarker.Rmd
update analysis

fit_model <- function(df_phenotype_feature, df_confounders_outcome) {
  df_model <- prepare_Xy(df_phenotype_feature, df_confounders_outcome)
  model_coeff <- data.frame(term = character(), estimate = double(), std.error = double(), statistic = double(), p.value = double())
  tryCatch({model_coeff = glm_Xy(df_model)}, error = function(e) print('cannot fit model'))
  #error = function(e) print('model did not run'))
  return (model_coeff)
}

@@

toc - tic
```
```{r}
## how many biomarkers we have screened
## at each time window
screen_results %>% group_by(lab_date_rel_bracket) %>% summarise(n = n_distinct(termid))
```

@@

screen_results_annotated <- screen_results %>%  filter(str_starts(term, 'is_testedTRUE')) %>% arrange(lab_date_rel_bracket) %>%
  left_join(hpo_term_list, by = 'termid') %>%
  select(lab_date_rel_bracket, termid, label, term, estimate, std.error, statistic, p.value)

@@

```{r}
# rank results by weight of is_tested
screen_results_annotated %>% group_by(lab_date_rel_bracket, termid) %>%
  arrange(term) %>%
  mutate(term_rank = row_number()) %>%
  ungroup() %>% 
  arrange(lab_date_rel_bracket, termid) %>% 
  group_by(lab_date_rel_bracket, termid, label) %>%
  nest() %>% 
  mutate(ranking_metric_value = map_dbl(data, function(df) df$estimate[df$term == 'is_testedTRUE'])) %>% 
  arrange(-ranking_metric_value) %>% 
  mutate(overall_rank = row_number()) %>%
  unnest(cols = c(data)) %>%
  write.csv(file = '../cache/screen_results_annotated_rank_by_weight_of_is_tested.csv', row.names = FALSE)

# rank results by weight of is_testedTRUE:is_observedTRUE
screen_results_annotated %>% group_by(lab_date_rel_bracket, termid) %>%
  arrange(term) %>%
  mutate(term_rank = row_number()) %>%
  ungroup() %>% 
  arrange(lab_date_rel_bracket, termid) %>% 
  group_by(lab_date_rel_bracket, termid, label) %>%
  nest() %>% 
  mutate(ranking_metric_value = map_dbl(data, function(df) df$estimate[df$term == 'is_testedTRUE:is_observedTRUE'])) %>% 
  arrange(-ranking_metric_value) %>% 
  mutate(overall_rank = row_number()) %>%
  unnest(cols = c(data)) %>%
  write.csv(file = '../cache/screen_results_annotated_rank_by_weight_of_is_tested_is_observed.csv', row.names = FALSE)

# rank results by p value of is_tested
screen_results_annotated %>% group_by(lab_date_rel_bracket, termid) %>%
  arrange(term) %>%
  mutate(term_rank = row_number()) %>%
  ungroup() %>% 
  arrange(lab_date_rel_bracket, termid) %>% 
  group_by(lab_date_rel_bracket, termid, label) %>%
  nest() %>% 
  mutate(ranking_metric_value = map_dbl(data, function(df) df$p.value[df$term == 'is_testedTRUE'])) %>% 
  arrange(ranking_metric_value) %>% 
  mutate(overall_rank = row_number()) %>%
  unnest(cols = c(data)) %>%
  write.csv(file = '../cache/screen_results_annotated_rank_by_p_of_is_tested.csv', row.names = FALSE)

# rank results by p value of is_testedTRUE:is_observedTRUE
screen_results_annotated %>% group_by(lab_date_rel_bracket, termid) %>%
  arrange(term) %>%
  mutate(term_rank = row_number()) %>%
  ungroup() %>% 
  arrange(lab_date_rel_bracket, termid) %>% 
  group_by(lab_date_rel_bracket, termid, label) %>%
  nest() %>% 
  mutate(ranking_metric_value = map_dbl(data, function(df) df$p.value[df$term == 'is_testedTRUE:is_observedTRUE'])) %>% 
  arrange(ranking_metric_value) %>% 
  mutate(overall_rank = row_number()) %>%
  unnest(cols = c(data)) %>%
  write.csv(file = '../cache/screen_results_annotated_rank_by_p_of_is_tested_is_observed.csv', row.names = FALSE)
```

>>>>

fit_model <- function(df_phenotype_feature, df_confounders_outcome) {
  df_model <- prepare_Xy(df_phenotype_feature, df_confounders_outcome)
  model_coeff <- data.frame(term = character(), estimate = double(), std.error = double(), statistic = double(), p.value = double())
  tryCatch({model_coeff = glm_Xy(df_model)}, error = function(e) warning('cannot fit model'))
  #error = function(e) print('model did not run'))
  return (model_coeff)
}

@@

toc - tic
```
```{r}
## how many biomarkers we have screened
## at each time window
N_BIOMARKERS_SCREENED <- screen_results %>% select(lab_date_rel_bracket, termid) %>% distinct() %>% nrow()
screen_results %>% group_by(lab_date_rel_bracket) %>% summarise(n = n_distinct(termid))
```

@@

screen_results_annotated <- screen_results %>%  filter(str_starts(term, 'is_testedTRUE')) %>% arrange(lab_date_rel_bracket) %>%
  left_join(hpo_term_list, by = 'termid') %>%
  select(lab_date_rel_bracket, termid, label, distance_to_root, term, distance_to_root, estimate, std.error, statistic, p.value)

@@

```{r}
screen_results_annotated_formatted <- screen_results_annotated %>% 
  mutate(OR = exp(estimate)) %>% 
  mutate(OR = round(OR, 2)) %>% 
  #mutate(p.value = scales::scientific(p.value, digits = 3)) %>% 
  select(-estimate, -std.error, -statistic) %>% 
  mutate(term = case_when(term == 'is_testedTRUE' ~ 'tested',
                          term == 'is_testedTRUE:is_observedTRUE' ~ 'observed',
                          TRUE ~ 'NEVER GONNA HAPPEN')) %>% 
  pivot_wider(id_cols = c(lab_date_rel_bracket, termid, label, distance_to_root), names_from = term, values_from = c(OR, p.value)) %>% 
  relocate(p.value_tested, .after = 'OR_tested') %>% 
  rename(`Odds Ratio (Been Tested)` = OR_tested) %>% 
  rename(`p value (Been Tested)` = p.value_tested) %>% 
  rename(`Odds Ratio (Been Observed)` = 'OR_observed') %>% 
  rename(`p value (Been Observed)` = 'p.value_observed')

write.csv(screen_results_annotated_formatted, file = '../cache/screen_results_annotated.csv', row.names = FALSE)

screen_results_annotated_formatted %>% head()
```

Highlight some biomarkers:

Eosinophilia
```{r}
eosinophilia <- screen_results_annotated_formatted %>% 
  filter(label == 'Eosinophilia') %>% 
  select(-distance_to_root) %>% 
  mutate(across(starts_with('p value'), scales::scientific))

sjPlot::tab_df(eosinophilia %>% rename(`Lab Year Bracket` = lab_date_rel_bracket), 
                title = 'Table Association of Eosinophilia with Severe Asthma',
               file = '../cache/table_eosinophilia_with_severe_asthma.html',
               alternate.rows = TRUE)

eosinophilia
```

Rank by Odds Ratio (Been Observed)
```{r}
significant_by_tested_and_observed <- screen_results_annotated_formatted %>% 
  arrange(desc(`Odds Ratio (Been Observed)`)) %>% 
  filter(`p value (Been Tested)` < 0.05/N_BIOMARKERS_SCREENED) %>%
  filter(`p value (Been Observed)` < 0.05/N_BIOMARKERS_SCREENED) %>%
  filter(`Odds Ratio (Been Tested)` > 1) %>% 
  filter(`Odds Ratio (Been Observed)` > 1) %>% 
  filter(lab_date_rel_bracket == '(-1,0]') %>%
  mutate(across(starts_with('p value'), scales::scientific))

significant_by_tested_and_observed_deduped <- significant_by_tested_and_observed %>% 
  rows_delete(tibble(label = c('Abnormal eosinophil morphology', 
                               'Abnormal leukocyte morphology',
                               'Abnormal immune system morphology',
                               'Abnormal cellular immune system morphology',
                               'Abnormal leukocyte count',
                               'Abnormal granulocyte morphology',
                               'Abnormality of the respiratory system',
                               'Functional respiratory abnormality')), by = 'label')

significant_by_tested_and_observed_deduped
```
```{r}
significant_by_tested_and_observed_deduped_three_years <- screen_results_annotated_formatted %>% 
  filter(termid %in% significant_by_tested_and_observed_deduped$termid) %>% 
  mutate(termid = factor(termid, levels = significant_by_tested_and_observed_deduped$termid)) %>% 
  arrange(termid, lab_date_rel_bracket) %>%
  mutate(across(starts_with('p value'), scales::scientific))

significant_by_tested_and_observed_deduped_three_years
```

```{r}
sjPlot::tab_df(significant_by_tested_and_observed_deduped %>% select(-distance_to_root, -lab_date_rel_bracket), 
                title = 'Table Selected Phenotypes (within 1 year prior) Associated with Severe Asthma',
               file = '../cache/table_selected_biomarker_from_labs_within_1_year_prior_to_dx.html', 
               alternate.rows = TRUE)
sjPlot::tab_df(significant_by_tested_and_observed_deduped_three_years %>% 
                 select(-distance_to_root) %>% rename(`Lab Year Bracket` = lab_date_rel_bracket), 
                title = 'Table Selected Phenotypes (within past 3 years) Associated with Severe Asthma',
               file = '../cache/table_selected_biomarker_from_labs_within_3_years_prior_to_dx.html',
               alternate.rows = TRUE)
```

### 2022-04-12 src/main/r/numeric_vs_hpo_for_liver_function.Rmd
update analysis

# cache directory to read from or write to
CACHE_DIR <- './cache'
if (!dir.exists(CACHE_DIR)){
  dir.create(CACHE_DIR)
}
```

@@

Retrieve liver function diagnoses information for the cohort.
```{r}
sql <- "SELECT * FROM cohort"
df_cohort <- get_df_from_db_or_cache(cache_dir = CACHE_DIR, file_name = 'liver_diagnoses.csv', db_access_callback = dbGetQuery, conn = dsca, statement = sql, force_rerun = FALSE)

@@

Retrieve liver function lab tests for the cohort. 
```{r}
sql <- read_file('../sql/get_labs_for_liver_function.sql')
df_liver_labs <- get_df_from_db_or_cache(cache_dir = CACHE_DIR, file_name = "labs_liver_function.csv", db_access_callback = dbGetQuery, conn=dsca, statement = sql, force_rerun = FALSE)
df_liver_labs %>% nrow()
```

@@

```{r}
require(furrr)
plan(multisession, workers = 4)

liver_analysis <- df_liver_labs_and_outcomes %>%
  group_by(loinc) %>%
  nest() %>%
  mutate(result = future_map(data, pipeline_compare_for_liver_function)) %>%
  select(-data) %>%
  unnest(result)

@@

lr_H <- lr_for_hpo_H(df_train = df_train, df_test = df_test)
roc_comparison <- compare_rocs(lr_H[[5]], lr_H[[6]])
roc_comparison <- bind_cols(roc_comparison, valid_lab_outcomes %>% filter(interpretation == 'H'))
roc_comparisons$H <- roc_comparison

>>>>

# cache directory to read from or write to
CACHE_DIR <- './cache'
if (!dir.exists(CACHE_DIR)){
  dir.create(CACHE_DIR)
}
FORCE_RERUN <- FALSE
```

@@

Retrieve liver function diagnoses information for the cohort.
```{r}
sql <- "SELECT * FROM cohort"
df_cohort <- get_df_from_db_or_cache(cache_dir = CACHE_DIR, file_name = 'liver_diagnoses.csv', db_access_callback = dbGetQuery, conn = dsca, statement = sql, force_rerun = FORCE_RERUN)

@@

Retrieve liver function lab tests for the cohort. 
```{r}
sql <- read_file('../sql/get_labs_for_liver_function.sql')
df_liver_labs <- get_df_from_db_or_cache(cache_dir = CACHE_DIR, file_name = "labs_liver_function.csv", db_access_callback = dbGetQuery, conn=dsca, statement = sql, force_rerun = FORCE_RERUN)
df_liver_labs %>% nrow()
```

@@

```{r}
require(furrr)
plan(multisession, workers = 6)

liver_analysis <- df_liver_labs_and_outcomes %>%
  group_by(loinc) %>%
  nest() %>%
  ungroup() %>% 
  mutate(result = future_map(data, pipeline_compare_for_liver_function)) %>%
  select(-data) %>%
  unnest(result)

write.csv(liver_analysis, file = 'cache/liver_analysis.csv', row.names = FALSE)

@@

lr_H <- lr_for_hpo_H(df_train = df_train, df_test = df_test)
roc_comparison <- compare_rocs(lr_H[[5]], lr_H[[6]])
